Variance reduction for Markov chains via martingale representations

### 2020-05-08 aap/Revision/vr-mcmc-revision-final.tex
update

\section{Covariance estimation for $V-$geometrically ergodic Markov chains}\label{sec:appendix_moments}
In this section we assume that $(X_p)_{p \in \nset_0}$ is a $V-$geometrically ergodic Markov chain and prove bounds on the variance of ergodic average $\pi_n^N(f)$ of the form \eqref{eq:29032018a2}. We use the same technique as in \cite{belomestny2019esvm} to control autocovariances for a given Markov chain. We start from some auxiliary lemmas:
\label{lem:fp_covariance}
Let $(X_p)_{p\geq 0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and let $f$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Set $\tilde{f}(x) = f(x) - \pi(f),$ then it holds for some constant $C > 0,$
|\mathsf{E}_x[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq 2C\rho^{s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}V(x),
|\mathsf{E}_{\pi}[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq 2C\pi(V)\rho^{s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}.
\label{lem:covariance}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and let $f$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Set $\tilde{f}(x) = f(x) - \pi(f),$ then
\label{eq:cov_v_ergodic}
\PCov_x \left[f(X_k),f(X_{k+s})\right] \leq C^2V^2(x)\rho^{2k+s} + C^2\rho^{k+s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x) + C\pi(V)\rho^{s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}
Note that
\PCov_x \left[f(X_k),f(X_{k+s})\right] &=& \PE_x\left[f(X_k) - \pi(f)\right]\left[f(X_{k+s})-\pi(f)\right] 
&&+ \left[\pi(f) - \PE_xf(X_{k+s})\right]\left[\PE_xf(X_k) - \pi(f)\right].
Due to $V-$ergodicity \eqref{eq:v_ergodicity},
\left|\left[\pi(f) - \PE_xf(X_{k+s})\right]\left[\PE_xf(X_k) - \pi(f)\right]\right| \leq C^2V^2(x)\rho^{2k+s}.
To bound the first term note that
\left|\PE_x\left[f(X_k) - \pi(f)\right]\left[f(X_{k+s})-\pi(f)\right] - \PCov_{\pi}\left[f(X_k),f(X_{k+s})\right]\right|
\leq \int\limits_{\rset^d}\left|\PE_y\left[\tilde{f}(X_0)\tilde{f}(X_{s})\right]\right| |\P^k(x,dy)-\pi(dy)|
\leq 2C^2\rho^{s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}\|\P^k(x,dy)-\pi(dy)\|_{V}
\leq 2C^2\rho^{k+s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}V(x),
where we have used Lemma~\ref{lem:fp_covariance} and ergodicity of $(X_k)_{k\geq 0}.$ This implies \eqref{eq:cov_v_ergodic}.
Now we state and prove the main result of this section on the variance bound for the estimator $\pi_N^n(f)$ in the case of a $V-$geometrically ergodic Markov chain.
\label{lem:var_bound}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and let $f$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$.  Then
\label{eq:vanilla_var}
\PVar_x \left[ \pi_{N}^n(f) \right] \leq \frac{C_1\pi(V)\|\tilde{f}\|^2_{V^{1/2}}}{n(1-\rho^{1/2})} + C_2\left(\frac{V(x) + 1}{n(1-\rho)}\right)^2
with constants $C_1,C_2 > 0$ and $\rho\in (0,1)$ from \eqref{eq:v_ergodicity}.
Note that
\PVar_x \left[ \frac{1}{n}\sum\limits_{k=N+1}^{N+n}f(X_k) \right] = \underbrace{\frac{1}{n^2}\sum\limits_{k=N+1}^{N+n}\PVar_x \left[ f(X_k) \right]}_{S_1} + \underbrace{\frac{2}{n^2}\sum\limits_{k=N+1}^{N+n-1}\sum\limits_{s=1}^{n-k-1}\PCov_x\left[f(X_k),f(X_{k+s})\right]}_{S_2}.
Now we bound each summand using Lemma~\ref{lem:covariance}:
S_1 \leq \frac{1}{n}C\pi(V)\|\tilde{f}\|_{V^{1/2}} + \frac{\rho^N C^2\left(V^2(x) + V(x)\|\tilde{f}\|_{V^{1/2}}\right)}{n^2(1-\rho)},
S_2 & \leq & \frac{2}{n^2}\sum\limits_{k=N+1}^{N+n-1}\left[\frac{C^2V^2(x)\rho^{2k+1}}{1-\rho} + \frac{2C^2V(x)\|\tilde{f}\|^2_{V^{1/2}}\rho^{k+1/2}}{1-\rho^{1/2}} + \frac{2C\|\tilde{f}\|^2_{V^{1/2}}\pi(V)}{1-\rho^{1/2}}\right] \leq \\
& \leq & \frac{2C^2\rho^{2N}V^2(x)}{n^2(1-\rho)^2} + \frac{4C^2\rho^{N+1/2}\|\tilde{f}\|^2_{V^{1/2}}V(x)}{n^2(1-\rho)(1-\rho^{1/2})} + \frac{4C\|\tilde{f}\|^2_{V^{1/2}}\pi(V)}{n(1-\rho^{1/2})}
Hence \eqref{eq:vanilla_var} follows.

@@

\begin{lem}
\label{lem:covariance}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Let $\tilde{f}(x) = f(x) - \pi(f)$. Then
\begin{equation}
\label{eq:cov_v_ergodic}
\PCov_x \left[f(X_k),f(X_{k+s})\right] \leq C^2V^2(x)\rho^{k+s} + C^2\rho^{k+s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x) + C\pi(V)\rho^{s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}
\end{equation}
\end{lem}
\begin{proof}
Note that
\[
\PCov_x \left[f(X_k),f(X_{k+s})\right] = \PE_x\left[f(X_k) - \pi(f)\right]\left[f(X_{k+s})-\pi(f)\right] + \left[\pi(f) - \PE_xf(X_s)\right]\left[\PE_xf(X_k) - \pi(f)\right]
\]
Due to $V-$ergodicity,
\[
\left|\left[\pi(f) - \PE_xf(X_s)\right]\left[\PE_xf(X_k) - \pi(f)\right]\right| \leq C^2V^2(x)\rho^{k+s}
\]
To bound the first term note that
\begin{multline*}
\left|\PE_x\left[f(X_k) - \pi(f)\right]\left[f(X_{k+s})-\pi(f)\right] - \PCov_{\pi}\left[f(X_k),f(X_{k+s})\right]\right|
\\
\leq \int\limits_{\rset^d}\left|\PE_y\left[\tilde{f}(X_0)\tilde{f}(X_{s})\right]\right| |\P^k(x,dy)-\pi(dy)|
\leq C^2\rho^{k+s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x),
\end{multline*}
where the last inequality is due to lemma~\ref{lem:fp_covariance}, which implies \eqref{eq:cov_v_ergodic}.
\end{proof}
Now we state and prove the main result of this section on the variance bound for the estimator $\pi_N^n(f)$ for $V-$geometrically ergodic Markov chain.
\begin{lem}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Assume also that $n(1-\rho^{1/2}) > 1$. Then
\begin{equation}
\label{eq:vanilla_var}
\PVar_x \left[ \pi_{N}^n(f) \right] \leq \frac{C\pi(V)\|\tilde{f}\|_{V^{1/2}}}{n(1-\rho^{1/2})} + \mathcal{O}(\frac{1}{n^2(1-\rho^{1/2})})
\end{equation}
\end{lem}
\begin{proof}
Note that
\begin{eqnarray*}
\PVar_x \left[ \frac{1}{n}\sum\limits_{k=N+1}^{N+n}f(X_k) \right] = \underbrace{\frac{1}{n^2}\sum\limits_{k=N+1}^{N+n}\PVar_x \left[ f(X_k) \right]}_{S_1} + \underbrace{\frac{2}{n^2}\sum\limits_{k=N+1}^{N+n-1}\sum\limits_{s=1}^{n-k-1}\PCov_x\left[f(X_k),f(X_{k+s})\right]}_{S_2}
\end{eqnarray*}
Now we bound first and second sum using lemma~\ref{lem:covariance}:
\[
S_1 \leq \frac{1}{n}C\pi(V)\|\tilde{f}\|_{V^{1/2}} + \frac{\rho^N C^2\left(V^2(x) + V(x)\|\tilde{f}\|_{V^{1/2}}\right)}{n^2(1-\rho)}
\]
\begin{eqnarray*}
S_2 & \leq & \frac{2}{n^2}\sum\limits_{k=N+1}^{N+n-1}\left[C^2V^2(x)\rho^{k+1}\frac{1}{1-\rho} + C^2V(x)\rho^{k+1/2}\frac{1}{1-\rho^{1/2}} + C\pi(V)\frac{1}{1-\rho^{1/2}}\|\tilde{f}\|_{V^{1/2}}\right] \leq \\
& \leq & \frac{2\rho^NC^2V^2(x)}{n^2(1-\rho)^2} + \frac{2\rho^NC^2V(x)}{n^2(1-\rho)(1-\rho^{1/2})} + \frac{C\pi(V)\|\tilde{f}\|_{V^{1/2}}}{n(1-\rho^{1/2})}
\end{eqnarray*}
Hence \eqref{eq:vanilla_var} follows.
\end{proof}

>>>>

\section{Covariance estimation for $V-$geometrically ergodic Markov chains}\label{sec:appendix_moments}
In this section we assume that $(X_p)_{p \in \nset_0}$ is a $V-$geometrically ergodic Markov chain and prove bounds on the variance of ergodic average $\pi_n^N(f)$ of the form \eqref{eq:29032018a2}. We use the same technique as in \cite{belomestny2019esvm} to control autocovariances for a given Markov chain. We start from some auxiliary lemmas:
\begin{lem}
\label{lem:fp_covariance}
Let $(X_p)_{p\geq 0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and let $f$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Set $\tilde{f}(x) = f(x) - \pi(f),$ then it holds for some constant $C > 0,$
\begin{equation}
|\mathsf{E}_x[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq 2C\rho^{s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}V(x),
\end{equation}
\begin{equation}
|\mathsf{E}_{\pi}[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq 2C\pi(V)\rho^{s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}.
\end{equation}
\end{lem}
\begin{lem}
\label{lem:covariance}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and let $f$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Set $\tilde{f}(x) = f(x) - \pi(f),$ then
\begin{equation}
\label{eq:cov_v_ergodic}
\PCov_x \left[f(X_k),f(X_{k+s})\right] \leq C^2V^2(x)\rho^{2k+s} + C^2\rho^{k+s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x) + C\pi(V)\rho^{s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}
\end{equation}
\end{lem}
\begin{proof}
Note that
\begin{eqnarray*}
\PCov_x \left[f(X_k),f(X_{k+s})\right] &=& \PE_x\left[f(X_k) - \pi(f)\right]\left[f(X_{k+s})-\pi(f)\right] 
\\
&&+ \left[\pi(f) - \PE_xf(X_{k+s})\right]\left[\PE_xf(X_k) - \pi(f)\right].
\end{eqnarray*}
Due to $V-$ergodicity \eqref{eq:v_ergodicity},
\[
\left|\left[\pi(f) - \PE_xf(X_{k+s})\right]\left[\PE_xf(X_k) - \pi(f)\right]\right| \leq C^2V^2(x)\rho^{2k+s}.
\]
To bound the first term note that
\begin{multline*}
\left|\PE_x\left[f(X_k) - \pi(f)\right]\left[f(X_{k+s})-\pi(f)\right] - \PCov_{\pi}\left[f(X_k),f(X_{k+s})\right]\right|
\\
\leq \int\limits_{\rset^d}\left|\PE_y\left[\tilde{f}(X_0)\tilde{f}(X_{s})\right]\right| |\P^k(x,dy)-\pi(dy)|
\\
\leq 2C^2\rho^{s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}\|\P^k(x,dy)-\pi(dy)\|_{V}
\leq 2C^2\rho^{k+s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}V(x),
\end{multline*}
where we have used Lemma~\ref{lem:fp_covariance} and ergodicity of $(X_k)_{k\geq 0}.$ This implies \eqref{eq:cov_v_ergodic}.
\end{proof}
Now we state and prove the main result of this section on the variance bound for the estimator $\pi_N^n(f)$ in the case of a $V-$geometrically ergodic Markov chain.

\begin{lem}
\label{lem:var_bound}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and let $f$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Denote $\tilde{f}(x) = f(x) - \pi(f)$. Then
\begin{equation}
\label{eq:vanilla_var}
\PVar_x \left[ \pi_{N}^n(f) \right] \leq \frac{C_1\pi(V)\|\tilde{f}\|^2_{V^{1/2}}}{n(1-\rho^{1/2})} + C_2\left(\frac{V(x) + 1}{n(1-\rho)}\right)^2
\end{equation}
with constants $C_1,C_2 > 0$ and $\rho\in (0,1)$ from \eqref{eq:v_ergodicity}.
\end{lem}
\begin{proof}
Note that
\begin{eqnarray*}
\PVar_x \left[ \frac{1}{n}\sum\limits_{k=N+1}^{N+n}f(X_k) \right] = \underbrace{\frac{1}{n^2}\sum\limits_{k=N+1}^{N+n}\PVar_x \left[ f(X_k) \right]}_{S_1} + \underbrace{\frac{2}{n^2}\sum\limits_{k=N+1}^{N+n-1}\sum\limits_{s=1}^{n-k-1}\PCov_x\left[f(X_k),f(X_{k+s})\right]}_{S_2}.
\end{eqnarray*}
Now we bound each summand using Lemma~\ref{lem:covariance}:
\[
S_1 \leq \frac{1}{n}C\pi(V)\|\tilde{f}\|_{V^{1/2}} + \frac{\rho^N C^2\left(V^2(x) + V(x)\|\tilde{f}\|_{V^{1/2}}\right)}{n^2(1-\rho)},
\]
\begin{eqnarray*}
S_2 & \leq & \frac{2}{n^2}\sum\limits_{k=N+1}^{N+n-1}\left[\frac{C^2V^2(x)\rho^{2k+1}}{1-\rho} + \frac{2C^2V(x)\|\tilde{f}\|^2_{V^{1/2}}\rho^{k+1/2}}{1-\rho^{1/2}} + \frac{2C\|\tilde{f}\|^2_{V^{1/2}}\pi(V)}{1-\rho^{1/2}}\right] \leq \\
& \leq & \frac{2C^2\rho^{2N}V^2(x)}{n^2(1-\rho)^2} + \frac{4C^2\rho^{N+1/2}\|\tilde{f}\|^2_{V^{1/2}}V(x)}{n^2(1-\rho)(1-\rho^{1/2})} + \frac{4C\|\tilde{f}\|^2_{V^{1/2}}\pi(V)}{n(1-\rho^{1/2})}
\end{eqnarray*}
Hence \eqref{eq:vanilla_var} follows.
\end{proof}

@@

\begin{lem}
\label{lem:covariance}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Denote $\tilde{f}(x) = f(x) - \pi(f)$. Then
\begin{equation}
\label{eq:cov_v_ergodic}
\PCov_x \left[f(X_k),f(X_{k+s})\right] \leq C^2V^2(x)\rho^{k+s} + C^2\rho^{k+s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x) + C\pi(V)\rho^{s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}
\end{equation}
\end{lem}
\begin{proof}
Note that
\[
\PCov_x \left[f(X_k),f(X_{k+s})\right] = \PE_x\left[f(X_k) - \pi(f)\right]\left[f(X_{k+s})-\pi(f)\right] + \left[\pi(f) - \PE_xf(X_s)\right]\left[\PE_xf(X_k) - \pi(f)\right]
\]
Due to $V-$ergodicity,
\[
\left|\left[\pi(f) - \PE_xf(X_s)\right]\left[\PE_xf(X_k) - \pi(f)\right]\right| \leq C^2V^2(x)\rho^{k+s}
\]
To bound the first term note that
\begin{multline*}
\left|\PE_x\left[f(X_k) - \pi(f)\right]\left[f(X_{k+s})-\pi(f)\right] - \PCov_{\pi}\left[f(X_k),f(X_{k+s})\right]\right|
\\
\leq \int\limits_{\rset^d}\left|\PE_y\left[\tilde{f}(X_0)\tilde{f}(X_{s})\right]\right| |\P^k(x,dy)-\pi(dy)|
\leq C^2\rho^{k+s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x),
\end{multline*}
where the last inequality is due to lemma~\ref{lem:fp_covariance}, which implies \eqref{eq:cov_v_ergodic}.
\end{proof}
Now we state and prove the main result of this section on the variance bound for the estimator $\pi_N^n(f)$ for $V-$geometrically ergodic Markov chain.
\begin{lem}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Assume also that $n(1-\rho^{1/2}) > 1$. Then
\begin{equation}
\label{eq:vanilla_var}
\PVar_x \left[ \pi_{N}^n(f) \right] \leq \frac{C\pi(V)\|\tilde{f}\|_{V^{1/2}}}{n(1-\rho^{1/2})} + \mathcal{O}(\frac{1}{n^2(1-\rho^{1/2})})
\end{equation}
\end{lem}
\begin{proof}
Note that
\begin{eqnarray*}
\PVar_x \left[ \frac{1}{n}\sum\limits_{k=N+1}^{N+n}f(X_k) \right] = \underbrace{\frac{1}{n^2}\sum\limits_{k=N+1}^{N+n}\PVar_x \left[ f(X_k) \right]}_{S_1} + \underbrace{\frac{2}{n^2}\sum\limits_{k=N+1}^{N+n-1}\sum\limits_{s=1}^{n-k-1}\PCov_x\left[f(X_k),f(X_{k+s})\right]}_{S_2}
\end{eqnarray*}
Now we bound first and second sum using lemma~\ref{lem:covariance}:
\[
S_1 \leq \frac{1}{n}C\pi(V)\|\tilde{f}\|_{V^{1/2}} + \frac{\rho^N C^2\left(V^2(x) + V(x)\|\tilde{f}\|_{V^{1/2}}\right)}{n^2(1-\rho)}
\]
\begin{eqnarray*}
S_2 & \leq & \frac{2}{n^2}\sum\limits_{k=N+1}^{N+n-1}\left[C^2V^2(x)\rho^{k+1}\frac{1}{1-\rho} + C^2V(x)\rho^{k+1/2}\frac{1}{1-\rho^{1/2}} + C\pi(V)\frac{1}{1-\rho^{1/2}}\|\tilde{f}\|_{V^{1/2}}\right] \leq \\
& \leq & \frac{2\rho^NC^2V^2(x)}{n^2(1-\rho)^2} + \frac{2\rho^NC^2V(x)}{n^2(1-\rho)(1-\rho^{1/2})} + \frac{C\pi(V)\|\tilde{f}\|_{V^{1/2}}}{n(1-\rho^{1/2})}
\end{eqnarray*}
Hence \eqref{eq:vanilla_var} follows.
\end{proof}
